# Bringing Data Analysis to the Sports Section

## NICAR 2024

#### Matt Waite, University of Nebraska

#### Derek Willis, University of Maryland

#### <https://github.com/dwillis/nicar24-sports>

Some of the best sports data stories start with a good question. How about this one: which teams benefit the most from home-court officiating?

First, we'll load the tidyverse.

```{r}
library(tidyverse)
```

```{r}
logs <- read_csv("logs1824.csv")

logs <- logs |> mutate(foul_diff = TeamPersonalFouls-OpponentPersonalFouls, steal_diff = TeamSteals- OpponentSteals, turnover_diff = TeamTurnovers- OpponentTurnovers)
```


```{r}

team_fouls_location_season <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, Team, Season, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

conference_fouls_location_season <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, Season, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

team_fouls_location <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, Team, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)


conference_fouls_location <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)
```
## Averages for all teams

```{r}
logs |> filter(!is.na(TeamPersonalFouls)) |> group_by(HomeAway) |> summarize(avg_fouls = mean(TeamPersonalFouls))
```

```{r}
team_fouldiff_location_season <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Conference, Team, Season, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

conference_fouldiff_location_season <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Conference, Season, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

team_fouldiff_location <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Team, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)


conference_fouldiff_location <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Conference, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)
```
### How many teams have averaged fewer fouls at home and away over the past seven seasons?

```{r}
team_fouldiff_location_season |> filter(Home < 0, Away < 0) |> group_by(Team) |> summarize(count = n()) |> arrange(desc(count))
```

### Teams that are good at this - are they alike? Yes, except South Carolina and Western Illinois.

```{r}
team_fouldiff_location |> 
  filter(Away < 0, Home < 0) |> 
  select(Team, Away, Home, Diff) |> 
  arrange(Diff)
```

### Drilling down into game-level data for West Virginia


```{r}
logs |> filter(Team == "West Virginia", !is.na(W_L), is.na(HomeAway)) |>  select(Season, Date, Opponent, W_L, foul_diff, steal_diff, turnover_diff) |>  arrange(desc(Date)) |> distinct()
```



