# Bringing Data Analysis to the Sports Section

## NICAR 2024

#### Matt Waite, University of Nebraska

#### Derek Willis, University of Maryland

#### <https://github.com/dwillis/nicar24-sports>

Some of the best sports data stories start with a good question. How about this one: which teams benefit the most from home-court officiating?

First, we'll load the tidyverse.

```{r}
library(tidyverse)
```

Our data is game-level records from the 2017-18 season through last week. After loading it, we'll want to add columns to measure the foul differential between the two teams, and I'll through in steal differential, too, since that's often associated with aggressive defense.

```{r}
logs <- read_csv("logs1824.csv")

logs <- logs |> mutate(foul_diff = TeamPersonalFouls-OpponentPersonalFouls, steal_diff = TeamSteals- OpponentSteals, turnover_diff = TeamTurnovers- OpponentTurnovers)
```

Let's take a look at our data so we can get familiar with it. 

## Averages for all teams

```{r}
logs |> filter(!is.na(TeamPersonalFouls)) |> group_by(HomeAway) |> summarize(avg_fouls = mean(TeamPersonalFouls))
```


Then, we'll start our group_by and summarize work, for teams and conferences, thanks to `pivot_wider`:

```{r}

team_fouls_location_season <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, Team, Season, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

conference_fouls_location_season <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, Season, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

team_fouls_location <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, Team, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)


conference_fouls_location <- logs |> 
  filter(!is.na(TeamPersonalFouls)) |> 
  group_by(Conference, HomeAway) |> 
  summarize(avg_fouls = mean(TeamPersonalFouls)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

team_fouldiff_location_season <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Conference, Team, Season, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

conference_fouldiff_location_season <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Conference, Season, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

team_fouldiff_location <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Team, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)

conference_fouldiff_location <- logs |> 
  filter(!is.na(foul_diff)) |> 
  group_by(Conference, HomeAway) |> 
  summarize(avg_fouls = mean(foul_diff)) |> 
  pivot_wider(names_from = HomeAway, values_from = avg_fouls) |> 
  rename(Home = `NA`, Away = `@`, Neutral = N) |> 
  mutate(Diff = Home-Away)
```
### How many teams have averaged fewer fouls at home and away over the past seven seasons?

```{r}
team_fouldiff_location_season |> filter(Home < 0, Away < 0) |> group_by(Team) |> summarize(count = n()) |> arrange(desc(count))
```

### Teams that are good at this - are they alike? Yes, except South Carolina and Western Illinois.

```{r}
team_fouldiff_location |> 
  filter(Away < 0, Home < 0) |> 
  select(Team, Away, Home, Diff) |> 
  arrange(Diff)
```

### Drilling down into game-level data for West Virginia


```{r}
logs |> filter(Team == "West Virginia", !is.na(W_L), is.na(HomeAway)) |>  select(Season, Date, Opponent, W_L, foul_diff, steal_diff, turnover_diff) |>  arrange(desc(Date)) |> distinct()
```


